---
title: "COMP2025 Introduction to Data Science — Assignment 1"
author: "Pham Ninh Khoi"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    number_sections: true
  html_document:
    toc: true
    number_sections: true
---

# Declaration

By submitting this work I certify that: no part of it has been copied from another student or any other source except where acknowledged; no part was written or produced for me by another person except as authorised by the lecturer; I have disclosed all use of generative-AI tools in the Reflection Log; and I understand this work may be checked by plagiarism- and AI-detection software.

Signed (typed name): Pham Ninh Khoi  
Student ID: 22145599  
Date: `r Sys.Date()`

# 0. Setup and Reproducibility

This section records the R environment, required packages, personalised seed, and data files used in this report. These details make the analysis reproducible, especially for random procedures such as cross-validation and train-test splitting.

**a) R environment and packages**

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

```{r packages}
library(boot)
library(pROC)
```

```{r r-environment}
R.version.string
packageVersion("boot")
packageVersion("pROC")
```

The output above records the R version and package versions used in this analysis. The `boot` package is used later for 10-fold cross-validation, while the `pROC` package is used later to produce the ROC curve and calculate AUC for the classification model.

**b) Personalised seed**

```{r personalised-seed}
D <- 599
set.seed(D)

D
```

The personalised seed value is **599**, which is the last three digits of Student ID **22145599**. The command `set.seed(599)` is used wherever a random procedure is required. In this assignment, the seed affects the 10-fold cross-validation in Question 1, the 80/20 train-test split in Question 2, and the 10-fold cross-validation in Question 2.

**c) Data files used**

```{r data-files-used}
energy_file <- "data/energy_homes.csv"
learners_file <- "data/learners.csv"

file.exists(energy_file)
file.exists(learners_file)
```

Both required data files were found successfully, as shown by the two `TRUE` outputs. The file `energy_homes.csv` is used for Question 1, which is the regression task predicting `heating_load`. The file `learners.csv` is used for Question 2, which is the classification task predicting whether a learner completed the course.

# 1. Question 1 — Regression

This question uses `energy_homes.csv` to model `heating_load`, which is the annual heating energy demand per unit floor area. The goal is to understand how building design variables are related to heating demand.

## 1.1 Exploration and Data Thinking

**a) Data import, dimensions, and missing values**

```{r q1-1-a-import-energy}
energy <- read.csv("data/energy_homes.csv")

dim(energy)
names(energy)
str(energy)
colSums(is.na(energy))
```

The `energy_homes.csv` dataset contains **768 rows** and **12 columns**. Each row represents one simulated residential building. The columns include one identifier variable, ten predictor variables, and one response variable, `heating_load`. The missing-value check shows that every column has **0 missing values**, so no rows need to be removed and no imputation is required.

```{r q1-1-a-variable-roles}
response <- "heating_load"

predictors <- c(
  "relative_compactness",
  "surface_area",
  "wall_area",
  "roof_area",
  "overall_height",
  "orientation",
  "glazing_area",
  "glazing_distribution",
  "wall_insul_rating",
  "window_age"
)

response
predictors
```

The response variable is `heating_load`, which measures annual heating demand in kWh/m². The variable `home_id` is only a building identifier, so it is not used as a predictor. The ten predictors used in this question are the building design variables listed in the output above.

**b) Scatterplot matrix, correlation matrix, and key relationships**

```{r q1-1-b-scatterplot-matrix, fig.height=10, fig.width=10}
energy_for_plots <- energy[, c(predictors, response)]

pairs(
  energy_for_plots,
  main = "Scatterplot Matrix of Heating Load and Predictors",
  pch = 19,
  cex = 0.18,
  cex.labels = 0.7
)
```

The scatterplot matrix provides an initial visual overview of the relationships between `heating_load` and the ten predictors. Because several predictors are coded or discrete variables, such as `orientation`, `overall_height`, `glazing_area`, and `glazing_distribution`, some panels appear as vertical or horizontal bands rather than smooth clouds of points. This is expected and does not indicate an error.

```{r q1-1-b-correlation-matrix}
cor_matrix <- cor(energy_for_plots)

round(cor_matrix, 3)
```

```{r q1-1-b-correlation-with-heating}
cor_with_heating <- cor_matrix["heating_load", predictors]

round(cor_with_heating, 3)

ranked_correlations <- sort(abs(cor_with_heating), decreasing = TRUE)

round(ranked_correlations, 3)
```

Based on the correlations with `heating_load`, the strongest linear relationship is with `overall_height`, with a correlation of approximately **0.647**. This positive correlation suggests that taller buildings tend to have higher heating load in this dataset. The second strongest relationship is with `relative_compactness`, with a correlation of approximately **-0.520**, meaning more compact buildings tend to have lower heating load. By contrast, `orientation` has a correlation of approximately **0.009**, which is almost zero, so it appears to have essentially no linear relationship with heating demand. `window_age` is also very weak, with a correlation of approximately **0.033**.

**c) Bias–variance discussion**

A model using only the strongest predictor, `overall_height`, would likely have **relatively high bias but low variance**. It would have low variance because a one-predictor model is simple and its predictions are less sensitive to small changes in the training data. However, it would likely have high bias because `heating_load` is also related to other predictors, especially `relative_compactness`, `surface_area`, and `wall_insul_rating`, so using only `overall_height` would ignore important building-design information.

## 1.2 Simple Linear and Polynomial Regression

**a) Simple linear regression model and estimated equation**

```{r q1-2-a-simple-linear-model}
simple_model <- lm(heating_load ~ relative_compactness, data = energy)

summary(simple_model)
```

```{r q1-2-a-simple-equation}
coef(simple_model)
```

A simple linear regression model was fitted to predict `heating_load` from `relative_compactness`. The estimated regression equation is:

**Predicted heating_load = 130.871 - 55.906 * relative_compactness**

This means that the predicted heating load decreases as `relative_compactness` increases. This model is used as a baseline because it uses only one predictor to explain heating demand.

**b) Hypothesis test and slope interpretation**

For the slope of `relative_compactness`, the hypotheses are:

H0: beta_1 = 0. There is no linear relationship between `relative_compactness` and `heating_load`.

H1: beta_1 is not equal to 0. There is a non-zero linear relationship between `relative_compactness` and `heating_load`.

```{r q1-2-b-slope-test}
summary(simple_model)$coefficients
```

The p-value for `relative_compactness` is approximately **1.96e-54**, which is far below 0.05. Therefore, the null hypothesis is rejected. There is strong statistical evidence that `relative_compactness` is a significant linear predictor of `heating_load`.

The slope for `relative_compactness` is approximately **-55.906**. This means that a one-unit increase in `relative_compactness` is associated with an estimated **55.906 kWh/m² decrease** in annual heating demand. Since a full one-unit increase is large on the compactness scale, a more practical interpretation is that a 0.1 increase in `relative_compactness` is associated with an estimated decrease of about **5.591 kWh/m²**.

**c) Standard error and 95% confidence interval**

```{r q1-2-c-standard-error-ci}
summary(simple_model)$coefficients["relative_compactness", "Std. Error"]

confint(simple_model)["relative_compactness", ]
```

The standard error of the slope is approximately **3.318**. The 95% confidence interval for the slope is approximately **[-62.419, -49.393]**.

This confidence interval does not include zero, which supports the conclusion that `relative_compactness` is statistically significant. The interval is reasonably narrow relative to the size of the slope, suggesting that the estimate is fairly precise.

**d) R², residual standard error, and adequacy for design decisions**

```{r q1-2-d-r2-rse}
summary(simple_model)$r.squared
summary(simple_model)$adj.r.squared
summary(simple_model)$sigma
```

The R-squared value is approximately **0.270**, meaning that `relative_compactness` alone explains about **27.0%** of the variation in `heating_load`.

The residual standard error is approximately **9.428 kWh/m²**, so predictions from this model are typically about 9.428 kWh/m² away from the actual heating load values. This single-predictor model is useful as a baseline, but it is probably not adequate for design decisions because around 73% of the variation remains unexplained.

**e) Diagnostic plots and model assumptions**

```{r q1-2-e-diagnostic-plots, fig.height=4.5, fig.width=12}
par(mfrow = c(1, 3))

plot(simple_model, which = 1)
plot(simple_model, which = 2)
plot(simple_model, which = 3)

par(mfrow = c(1, 1))
```

The residuals vs fitted plot shows some visible pattern rather than a completely random cloud, so the linearity assumption is not fully satisfied.

The Normal Q-Q plot shows that most residuals are reasonably close to the straight line, although there are some departures at the tails, so the normality assumption is broadly acceptable but not perfect.

The scale-location plot shows some variation in residual spread across fitted values, so the constant-variance assumption is not fully satisfied. Overall, the simple linear model is useful as a baseline, but it does not fully capture the structure in the data.

**f) Polynomial regression for `glazing_area` and model selection**

```{r q1-2-f-polynomial-models}
poly_1 <- lm(heating_load ~ poly(glazing_area, 1, raw = TRUE), data = energy)
poly_2 <- lm(heating_load ~ poly(glazing_area, 2, raw = TRUE), data = energy)
poly_3 <- lm(heating_load ~ poly(glazing_area, 3, raw = TRUE), data = energy)
poly_4 <- lm(heating_load ~ poly(glazing_area, 4, raw = TRUE), data = energy)
poly_5 <- lm(heating_load ~ poly(glazing_area, 5, raw = TRUE), data = energy)

poly_results <- data.frame(
  Order = c(1, 2, 3, 4, 5),
  Adjusted_R_Squared = c(
    summary(poly_1)$adj.r.squared,
    summary(poly_2)$adj.r.squared,
    summary(poly_3)$adj.r.squared,
    summary(poly_4)$adj.r.squared,
    summary(poly_5)$adj.r.squared
  ),
  AIC = c(
    AIC(poly_1),
    AIC(poly_2),
    AIC(poly_3),
    AIC(poly_4),
    AIC(poly_5)
  )
)

poly_results
```

Polynomial regression models of orders 1 to 5 were fitted to examine whether the relationship between `glazing_area` and `heating_load` is better represented by a curved relationship rather than a straight line. Adjusted R-squared and AIC are used because they consider both model fit and model complexity.

Based on the polynomial model comparison, the order 2 polynomial is preferred. The adjusted R-squared increases from approximately **0.021** for order 1 to approximately **0.058** for order 2, and AIC decreases from approximately **5854.8** to approximately **5826.1**. Higher-order models do not provide a meaningful improvement, because orders 3, 4, and 5 have slightly lower adjusted R-squared and higher AIC than the order 2 model. Therefore, order 2 gives the best balance between fit and unnecessary complexity.